---
title: "Homework 2"
author: "Zeheng Wei"
date: 2026-09-30
output: github_document
---

```{r setup, include = FALSE}
library(tidyverse)
```

# Problem 1

### Import and clean data

```{r import_and_clean_subway_entrance_exit_data, warning = FALSE}
nyc_subway_entrance_exit_data = read_csv(
  file = "./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
  col_types = cols(
    Route8 = col_character(),
    Route9 = col_character(),
    Route10 = col_character(),
    Route11 = col_character())) |>
  janitor::clean_names() |>
  select(
    line,
    station_name,
    station_latitude,
    station_longitude,
    route1:route11,
    entry,
    vending,
    entrance_type,
    ada) |>
  mutate(
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE))
```

### Check the cleaned data

```{r check_subway_entrance_exit_data}
nyc_subway_entrance_exit_data

number_of_entrance_exit_records =
  length(nyc_subway_entrance_exit_data$line)

number_of_retained_variables =
  length(names(nyc_subway_entrance_exit_data))
```

### Dataset description

Each row describes a NYC subway entrance or exit, so stations may appear more than once. Variables include the line, station name and location, routes, entry permission, vending, entrance type, and ADA compliance. I cleaned the column names, kept the requested variables, converted entry from YES/NO to TRUE/FALSE, and read Route8–Route11 as character columns for consistent route types. The cleaned data have `r number_of_entrance_exit_records` rows and `r number_of_retained_variables` columns. They are not tidy yet because routes are spread across 11 columns instead of one variable.

### Distinct stations

```{r count_distinct_subway_stations}
number_of_distinct_stations =
  nyc_subway_entrance_exit_data |>
  distinct(line, station_name) |>
  pull(station_name) |>
  length()
```

There are `r number_of_distinct_stations` distinct stations, identified using both line and station name.

### ADA-compliant stations

```{r count_ada_compliant_subway_stations}
number_of_ada_compliant_stations =
  nyc_subway_entrance_exit_data |>
  filter(ada == TRUE) |>
  distinct(line, station_name) |>
  pull(station_name) |>
  length()
```

For these stations, `r number_of_ada_compliant_stations` are ADA compliant.

### Entry without vending

```{r calculate_entry_proportion_without_vending}
number_of_entrances_exits_without_vending =
  nyc_subway_entrance_exit_data |>
  filter(vending == "NO") |>
  pull(entry) |>
  length()

number_of_entrances_exits_without_vending_allowing_entry =
  nyc_subway_entrance_exit_data |>
  filter(vending == "NO", entry == TRUE) |>
  pull(entry) |>
  length()

proportion_allowing_entry_without_vending =
  number_of_entrances_exits_without_vending_allowing_entry /
  number_of_entrances_exits_without_vending
```

Among the `r number_of_entrances_exits_without_vending` entrances and exits without vending, `r number_of_entrances_exits_without_vending_allowing_entry` allow entry, giving a proportion of `r round(proportion_allowing_entry_without_vending, digits = 2)`.

### Routes and A train stations

```{r reformat_routes_and_count_a_train_stations}
nyc_subway_route_data =
  nyc_subway_entrance_exit_data |>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name") |>
  drop_na(route_name)

number_of_a_train_stations =
  nyc_subway_route_data |>
  filter(route_name == "A") |>
  distinct(line, station_name) |>
  pull(station_name) |>
  length()

number_of_ada_compliant_a_train_stations =
  nyc_subway_route_data |>
  filter(route_name == "A", ada == TRUE) |>
  distinct(line, station_name) |>
  pull(station_name) |>
  length()
```

I reshaped the routes into `route_number` and `route_name` and removed empty slots. Using line and station name to identify distinct stations, the A train serves `r number_of_a_train_stations` stations, of which `r number_of_ada_compliant_a_train_stations` are ADA compliant.



# Problem 2

### Import and clean Mr. Trash Wheel data

```{r import_and_clean_mr_trash_wheel_data}
mr_trash_wheel_data =
  readxl::read_excel(
    "./data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Mr. Trash Wheel",
    range = "A2:N742") |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(
    year = as.integer(year),
    sports_balls = as.integer(round(sports_balls, digits = 0)),
    trash_wheel = "Mr. Trash Wheel")
```

### Import and clean Professor and Gwynnda data

```{r import_and_clean_professor_and_gwynnda_data}
professor_trash_wheel_data =
  readxl::read_excel(
    "./data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Professor Trash Wheel",
    range = "A2:M139",
    na = c("", "dive")) |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(
    year = as.integer(year),
    trash_wheel = "Professor Trash Wheel")

gwynnda_trash_wheel_data =
  readxl::read_excel(
    "./data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Gwynnda the Good Wheel of the W",
    range = "A2:L394") |>
  janitor::clean_names() |>
  drop_na(dumpster) |>
  mutate(
    year = as.integer(year),
    trash_wheel = "Gwynnda")
```

### Combine and check the data

```{r combine_and_check_trash_wheel_data}
trash_wheel_collection_data =
  bind_rows(
    mr_trash_wheel_data,
    professor_trash_wheel_data,
    gwynnda_trash_wheel_data)

trash_wheel_collection_data

number_of_dumpster_records =
  length(trash_wheel_collection_data$dumpster)
```

### Dataset description

The combined dataset contains `r number_of_dumpster_records` dumpster records from Mr. Trash Wheel, Professor Trash Wheel, and Gwynnda. Variables include collection date, weight in tons, volume in cubic yards, homes powered, and trash item counts. I excluded totals and non-data entries, cleaned column names, and kept rows with a dumpster number. I rounded Mr. Trash Wheel's sports ball counts and converted them to integers, standardized `year`, and added `trash_wheel` to identify the source. I treated the nonnumeric value `dive` in Professor's plastic bottle counts as missing. Missing or unrecorded measurements remain `NA`.

### Total weight collected

```{r calculate_total_weight}
total_weight_in_tons_collected_by_professor_trash_wheel =
  trash_wheel_collection_data |>
  filter(trash_wheel == "Professor Trash Wheel") |>
  pull(weight_tons) |>
  sum()
```

Across the available records, Professor Trash Wheel collected `r round(total_weight_in_tons_collected_by_professor_trash_wheel, digits = 2)` tons of trash.

### Cigarette butts collected by Gwynnda

```{r calculate_total_cigarette_butts}
total_cigarette_butts_collected_by_gwynnda_in_june_2022 =
  trash_wheel_collection_data |>
  filter(trash_wheel == "Gwynnda", year == 2022, month == "June") |>
  pull(cigarette_butts) |>
  sum()
```

Gwynnda collected `r as.integer(total_cigarette_butts_collected_by_gwynnda_in_june_2022)` cigarette butts in June 2022.

# Problem 3

### Import and clean baseline demographics

```{r import_and_clean_mci_baseline_data, message = FALSE, warning = FALSE}
mci_baseline_imported_data =
  read_csv(
    file = "./data/data_mci/MCI_baseline.csv",
    skip = 1,
    na = c(".", "NA", "")) |>
  janitor::clean_names() |>
  rename(baseline_age = current_age) |>
  mutate(
    sex = case_match(sex, 0 ~ "female", 1 ~ "male"),
    sex = as.factor(sex),
    apoe4 = case_match(apoe4, 0 ~ "non-carrier", 1 ~ "carrier"),
    apoe4 = as.factor(apoe4))

mci_baseline_data =
  mci_baseline_imported_data |>
  filter(is.na(age_at_onset) | age_at_onset > baseline_age)
```

### Baseline participant statistics

```{r calculate_mci_baseline_statistics}
mci_baseline_data

number_of_recruited_participants =
  length(mci_baseline_imported_data$id)

number_of_eligible_participants =
  length(mci_baseline_data$id)

number_of_excluded_participants =
  number_of_recruited_participants - number_of_eligible_participants

number_of_participants_who_developed_mci =
  sum(!is.na(mci_baseline_data$age_at_onset))

average_baseline_age =
  mean(mci_baseline_data$baseline_age)

number_of_female_participants =
  sum(mci_baseline_data$sex == "female")

number_of_female_apoe4_carriers =
  sum(mci_baseline_data$sex == "female" & mci_baseline_data$apoe4 == "carrier")

proportion_of_female_participants_with_apoe4 =
  number_of_female_apoe4_carriers / number_of_female_participants
```

### Baseline data description and results

I skipped the description row, cleaned column names, renamed `current_age` to `baseline_age`, and converted sex and APOE4 status to labeled factors. I treated `.` as missing onset age and retained these participants because they remained MCI free throughout follow-up. Each row represents one participant, with baseline demographics, education, and MCI onset age.

The original baseline file contains `r number_of_recruited_participants` participants. I excluded `r number_of_excluded_participants` whose MCI onset age was at or before their baseline age, leaving `r number_of_eligible_participants` eligible participants. Among this eligible group, `r number_of_participants_who_developed_mci` developed MCI during follow-up, and the average baseline age was `r round(average_baseline_age, digits = 2)` years. Of the `r number_of_female_participants` women, `r number_of_female_apoe4_carriers` were APOE4 carriers, giving a proportion of `r round(proportion_of_female_participants_with_apoe4, digits = 2)`.

### Import and tidy amyloid data

```{r import_and_tidy_mci_amyloid_data, message = FALSE, warning = FALSE}
mci_amyloid_data =
  read_csv(
    file = "./data/data_mci/mci_amyloid.csv",
    skip = 1,
    na = c(".", "NA", "Na", "")) |>
  janitor::clean_names() |>
  rename(id = study_id) |>
  pivot_longer(
    baseline:time_8,
    names_to = "years_since_baseline",
    values_to = "amyloid_42_40_ratio") |>
  mutate(
    years_since_baseline = case_match(
      years_since_baseline,
      "baseline" ~ 0,
      "time_2" ~ 2,
      "time_4" ~ 4,
      "time_6" ~ 6,
      "time_8" ~ 8))

mci_amyloid_data

number_of_amyloid_visit_records =
  length(mci_amyloid_data$id)

number_of_participants_in_amyloid_data =
  mci_amyloid_data |>
  distinct(id) |>
  pull(id) |>
  length()

number_of_missing_amyloid_ratios =
  sum(is.na(mci_amyloid_data$amyloid_42_40_ratio))
```

### Amyloid dataset description

I skipped the description row, cleaned column names, renamed `study_id` to `id`, and treated `NA` and `Na` as missing. Using `pivot_longer()`, I reshaped the visit columns into time (0, 2, 4, 6, and 8 years since baseline) and amyloid 42/40 ratio variables. The tidy dataset contains `r number_of_amyloid_visit_records` participant-visit records from `r number_of_participants_in_amyloid_data` participants, with `r number_of_missing_amyloid_ratios` missing ratios retained as `NA`.

### Check participants appearing in only one dataset

```{r check_unmatched_mci_participants}
participants_only_in_baseline_data =
  anti_join(mci_baseline_data, mci_amyloid_data, by = "id") |>
  distinct(id)

participants_only_in_amyloid_data =
  anti_join(mci_amyloid_data, mci_baseline_data, by = "id") |>
  distinct(id)

amyloid_participants_without_original_baseline_data =
  anti_join(mci_amyloid_data, mci_baseline_imported_data, by = "id") |>
  distinct(id)

number_of_participants_only_in_baseline_data =
  length(participants_only_in_baseline_data$id)

number_of_participants_only_in_amyloid_data =
  length(participants_only_in_amyloid_data$id)

number_of_amyloid_participants_without_original_baseline_data =
  length(amyloid_participants_without_original_baseline_data$id)

number_of_excluded_participants_in_amyloid_data =
  number_of_participants_only_in_amyloid_data -
  number_of_amyloid_participants_without_original_baseline_data

participants_only_in_baseline_data
print(participants_only_in_amyloid_data, n = 20)
```

### Participant matching findings

Comparing the amyloid data with the eligible baseline dataset, `r number_of_participants_only_in_baseline_data` participants appear only in baseline and `r number_of_participants_only_in_amyloid_data` appear only in amyloid. Of the amyloid-only group, `r number_of_excluded_participants_in_amyloid_data` were excluded earlier because they already had MCI at baseline. Comparing with the original baseline file shows that the remaining `r number_of_amyloid_participants_without_original_baseline_data` participants have no baseline record. The source files do not explain the remaining mismatches.

### Combine baseline and amyloid data

```{r combine_mci_baseline_and_amyloid_data}
mci_baseline_amyloid_data =
  inner_join(mci_baseline_data, mci_amyloid_data, by = "id")

mci_baseline_amyloid_data

number_of_participants_in_combined_mci_data =
  mci_baseline_amyloid_data |>
  distinct(id) |>
  pull(id) |>
  length()

number_of_visit_records_in_combined_mci_data =
  length(mci_baseline_amyloid_data$id)

number_of_variables_in_combined_mci_data =
  length(names(mci_baseline_amyloid_data))
```

### Combined dataset description

I joined the eligible baseline and amyloid datasets by `id`, keeping participants present in both. The result contains `r number_of_visit_records_in_combined_mci_data` participant-visit records from `r number_of_participants_in_combined_mci_data` participants and `r number_of_variables_in_combined_mci_data` variables. Each row is one participant at one visit, with baseline demographics, MCI onset age, years since baseline, and the amyloid ratio. Baseline information is repeated across visits, and missing amyloid ratios remain as `NA`.

### Export the combined data

```{r export_combined_mci_data}
write_csv(mci_baseline_amyloid_data, "./data/mci_baseline_amyloid_data.csv")
```
