---
title: "Homework 2"
author: "Zeheng Wei"
date: 2026-09-30
output: github_document
---

```{r setup, include = FALSE}
library(tidyverse)
```

# Problem 1

### Import and clean data

```{r import_and_clean_subway_entrance_exit_data, warning = FALSE}
nyc_subway_entrance_exit_data = read_csv(
  file = "./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
  col_types = cols(
    Route8 = col_character(),
    Route9 = col_character(),
    Route10 = col_character(),
    Route11 = col_character())) |>
  janitor::clean_names() |>
  select(
    line,
    station_name,
    station_latitude,
    station_longitude,
    route1:route11,
    entry,
    vending,
    entrance_type,
    ada) |>
  mutate(
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE))
```

### Check the cleaned data

```{r check_subway_entrance_exit_data}
nyc_subway_entrance_exit_data

number_of_entrance_exit_records =
  length(nyc_subway_entrance_exit_data$line)

number_of_retained_variables =
  length(names(nyc_subway_entrance_exit_data))
```

### Dataset description

Each row describes a NYC subway entrance or exit, so stations may appear more than once. Variables include the line, station name and location, routes, entry permission, vending, entrance type, and ADA compliance. I cleaned the column names, kept the requested variables, converted entry from YES/NO to TRUE/FALSE, and read Route8–Route11 as character columns for consistent route types. The cleaned data have `r number_of_entrance_exit_records` rows and `r number_of_retained_variables` columns. They are not tidy yet because routes are spread across 11 columns instead of one variable.

# Problem 2

# Problem 3
